TEMPERATURES AND METALLICITIES OF M DWARFS IN THE \apogee\ SURVEY

### 2019-08-01 paper/draft.tex
discussion revisions

For the physical parameter model, we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model that 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening. We note these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and have cross-validation labels biased by an average of $+65\,$K and $-0.08\,$dex, however removing them from the training sample does not significantly change the overall scatter and bias of the model (when removed the scatters are 79\,K/0.09\,dex and biases are +5\,K/+0.01\,dex). Hence we do not remove them from the training sample.

@@

The model we have developed does have limitations however.
For example, it delivers chi-squared values---goodness-of-fit measures---that
are large; the model is not technically an accurate description of the spectra,
especially when the spectra are observed at signal-to-noise levels above 100.
For another, it does not include some known physical and instrumental effects.
It does not include a term or terms for line broadening from rotation or
convection (for exmaple, \citealt{Behmard2019}).
It does not include any accounting for binarity or superposition of multiple
stellar spectra (as in, say, \citealt{ElBadry:2018}).
And it does not include any adjustments for instrumental variations, like for
example the small but significant variations of \apogee\ resolution with
spectrograph fiber number (as are included in \citealt{Ness2018}).

In addition to all this, the \apogee\ instrument was designed to be sensitive to more than
a dozen individual element abundances in stellar spectra!
So the M-dwarf spectra analyzed here contain individual element abundance
information that we have left on the table.
Exploitation of that information might require better models of M dwarfs than
we have at present.
However, this is an important goal for the future with these data.

While a detailed analysis of atmospheric model limitations is beyond the scope of this paper, 
our results provide an avenue to compare the FGK star metallicity scale 
to the less well-understood M dwarf metallicity scale.
These results find that atmospheric metallicities are systematically metal-poor biased
compared to \cannon-based metallicities trained on sources with metallicities callibrated
to those of FGK companions. 
At the high temperature end ($\teff > 3600$\,K), the ASPCAP metallicity bias is $\sim-0.15\,$dex with a
scatter of $\sim0.06\,$dex relative to \cannon\ metallicities, and increases to a bias of $\>-0.26\,$dex 
and scatter of $\>0.1\,$dex at the lower temperature end ($\teff < 3300$\,K) (Figure \ref{fig:met_comp}).
This metal-poor bias, while not explored to a great extent in this work, we suspect is due 
to the line lists of the models: 
an effect in which the optimizer of the pipeline may be lowering the metallicity of the fit 
to better reproduce the spectrum to compensate for the missing lines or opacities.
Further analysis would need to be done to determine if the metallicity bias
is found in other model grids (besides the ATLAS/MARCS models used by the ASPCAP pipeline), 
and if the effect is present at other wavelengths.

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in Figures \ref{fig:demo_teff}, \ref{fig:demo_feh}, and \ref{fig:west_derivative} and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
These Tables summarize spectral features in the \apogee\ bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will, in the end, lead to
new and improved physical models, which will, in turn, put \thecannon\ out of business. \\

@@

\begin{figure*}[]
	\begin{center}
	\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
	\end{center}
	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallcities. \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}
	\label{fig:demo_teff}
\end{figure*}

>>>>

For the physical parameter model, we trained \thecannon\ on 87 M dwarfs with a two-dimensional temperature/metallicity labels, to a precision of 77K/0.09dex as estimated by the cross validation scatter, similar to the original training sample uncertainties of 60K/0.08dex. We note for this model that 5 out of 87 sources show possible rotational line broadening identified by visual inspection (as indicated by the red circles in Figure~\ref{fig:mann_validation}), while the remaining sources show no obvious broadening. We note that these broadened sources have high $\chi^2$ values (those sources with $\chi^2 > 80,000$ in Figure \ref{fig:chi_dist}), and have cross-validation labels biased by an average of $+65\,$K and $-0.08\,$dex, however removing them from the training sample does not significantly change the overall scatter and bias of the model (when removed the scatters are 79\,K/0.09\,dex and biases are +5\,K/+0.01\,dex). Hence we do not remove them from the training sample.

@@

The model we have developed does have limitations however.
For example, it delivers chi-squared values---goodness-of-fit measures---that
are large; the model is not technically an accurate description of the spectra,
especially when the spectra are observed at signal-to-noise levels above 100.
The model does not include some known physical and instrumental effects.
It does not include terms for line broadening from rotation or
convection (for example, \citealt{Behmard2019}), nor
does it account for binarity or superposition of multiple
stellar spectra (as in, say, \citealt{ElBadry:2018}).
The model also does not include any adjustments for instrumental variations, like for
example the small but significant variations of \apogee\ resolution with
spectrograph fiber number (as included in \citealt{Ness2018}).

The \apogee\ instrument was designed to be sensitive to more than
a dozen individual element abundances in stellar spectra.
So the M-dwarf spectra analyzed here contain individual element abundance
information that we have ignored.
Exploitation of that information requires a better training set of M dwarfs than
we have at present, but is an important goal for the future with these data.

While a detailed analysis of atmospheric model limitations is beyond the scope of this paper, 
our results provide an avenue to compare the FGK star metallicity scale 
to the less well-understood M dwarf metallicity scale.
These results find that atmospheric metallicities are systematically metal-poor biased
compared to \cannon-based metallicities trained on sources with metallicities calibrated
to those of FGK companions. 
At the high temperature end ($\teff > 3600$\,K), the ASPCAP metallicity bias is $\sim-0.15\,$dex with a
scatter of $\sim0.06\,$dex relative to \cannon\ metallicities, and increases to a bias of $\>-0.26\,$dex 
and scatter of $\>0.1\,$dex at the lower temperature end ($\teff < 3300$\,K) (Figure \ref{fig:met_comp}).
We suspect that this metal-poor bias, while not explored to a great extent in this work, is due 
to the line lists of the models: 
an effect in which the optimizer of the pipeline may be lowering the metallicity of the fit 
to compensate for the missing lines or opacities.
Further analysis would need to be done to determine if the metallicity bias
is found in other model grids (besides the ATLAS/MARCS models used by the ASPCAP pipeline), 
and if the effect is present at other wavelengths.

Given that physics-based spectral models of M dwarfs have issues, one of the possible
future values of the data-driven model shown here is that it is highly interpretable:
It contains within it first and second derivatives of the spectral expectation with
respect to the atmospheric parameters.
We show some of these derivatives in Figures \ref{fig:demo_teff}, \ref{fig:demo_feh}, and \ref{fig:west_derivative} and deliver relevant data in
Table~\ref{table:teff_derivative} and Table~\ref{table:feh_derivative}.
These Tables summarize spectral features in the \apogee\ bandpass
that are found to be strong temperature and metallicity indicators.
In the long run, this is the primary value of data-driven models for astronomy:
To provide physical insights that drive physical understandings.
It is our hope that \thecannon, and models like it, will lead to
new and improved physical models which will, in turn, put \thecannon\ out of business. \\

@@

\begin{figure*}[]
	\begin{center}
	\includegraphics[width=16cm]{figures/demo_derivatives_teff3.png}
	\end{center}
	\caption{\textit{Top two panels of each plot:} Mann-trained model for varying temperatures, and similar metallicities. \textit{Third panel of each plot:} Derivative of \thecannon\ model with respect to temperature, taken at the median training temperature, T$_{\rm eff}=3463$K; error estimate computed using a jackknife statistic at each pixel is marked in red, making it possible to distinguish which features vary significantly with change in spectral type, and which are likely due to noise.}
	\label{fig:demo_teff}
\end{figure*}
